Starcode: sequence clustering based on all-pairs search

### 2014-11-07 doc/document.tex
Manuscript: Updated HC dataset name.

The basic properties of the datasets used for benchmarking are
summarized in Table~\ref{tab:biobench_properties}. Dataset 1
(SRR950457) has been pre-processed to extract the barcode and remove
the constant part of the reads. Only barcodes between 15 and 17
nucleotides were included in the file. Dataset 2 (SRRxxxxxx) consists
of raw Illumina single reads. These datasets differ by the read size,
the total read count and by the empirical cluster sizes. According to
the output of starcode, the largest clusters of dataset 1 contain
approximately 70,000 sequences, whereas dataset 2 contains 4 clusters
with more than 1 million sequences. Dataset 3 (SRR950477) has been
included to benchmark starcode against rainbow in paired-end
clustering mode.
\begin{table}[h]
\centering
\caption{Summary of the biological datasets used for benchmarking.
All the datasets are Illumina reads.}
\begin{tabular}{l|l c r}
    \textit{dataset}   & \textit{read count} &
    \textit{read length} & \textit{type} \\ \hline
    SRR950457 & 6,542,309    & 16     & single     \\
    SRRxxxxxx & 127,675,537  & 50     & single     \\
    SRR950477 & 2,460,226    & 50+50  & paired-end \\
\end{tabular}
\label{tab:biobench_properties}
\end{table}

@@

\begin{table}[h]
\centering
\caption{Running time (in seconds) of the software on three biological
datasets. Runs longer than 10 days were interrupted. A dash indicates
that the software cannot be used for this dataset.}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{SRR950457} &
      \textit{SRRxxxxxx} & \textit{SRR950477} \\ \hline
    starcode   & 5           & 2,898       & 44    \\
    seed       & -           & 60,374      & -     \\
    sidesort   & 4,055       & $>$ 10 days & -     \\
    rainbow    & -           & -           & 306   \\
    cd-hit-est & 170         & 512,591     & -     \\
\end{tabular}
\label{tab:biobench_runtime}
\end{table}

@@

\begin{table}[h]
\centering
\caption{Memory usage (in GB).}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{SRR950457} &
      \textit{SRRxxxxxx} & \textit{SRR950477} \\ \hline
    starcode   & 0.65 & 30.9 & 5.2 \\
    seed       & -    & 53.9 & -   \\
    sidesort   & 1.30 & 13.9 & -   \\
    rainbow    & -    & -    & 0.5 \\
    cd-hit-est & 0.80 & 28.5 & -   \\
\end{tabular}
\label{tab:biobench_memory}
\end{table}

@@

\section{Discussion and conclusion}
Starocde is a solid algorithm for sequence clustering based on
all-pairs matching. It achieves high precision, and on experimental
datasets it can be faster than popular heuristics. By design,
starcode is tailored to process high throughput sequencing data
on multi-core platforms with sufficient amount of memory. Due to its
superior precision and faster running time, it fills a gap among
available software, by allowing to take full advantage of middle
to high end hardware.

The speed and precision of starcode also makes it useful for other
clustering tasks, such as identifying enriched motifs in microbial
genomes and in experimental data. Here we have given two examples of such
applications. In the first, we recover a known enriched 12-mer in the
genome of \textit{Neisseria meningitidis}. In the second, we recover
the motif of the human RNA binding protein SRSF1 and notice that it
seems to consist of two halves separated by a linker. This hypothesis
is consistent with the fact that SRSF1 binds RNA through two
consecutive RNA Recognition Motifs (RRM) that are known to bind
3-4 nucleotides in a row\citep{pmid23253355}. The Levenshtein distance,
which incorporates insertions and deletions is more likely to capture
bi-partite binding motifs than position weight matrix representations.
The use of a clustering method to tackle this problem is unusual, but it
it illustrates the potential advantages of distance-based approaches.

>>>>

The basic properties of the datasets used for benchmarking are
summarized in Table~\ref{tab:biobench_properties}. Dataset 1
(SRR950457) has been pre-processed to extract the barcode and remove
the constant part of the reads. Only barcodes between 15 and 17
nucleotides were included in the file. Dataset 2 (PRJEB7686) consists
of raw Illumina single reads. These datasets differ by the read size,
the total read count and by the empirical cluster sizes. According to
the output of starcode, the largest clusters of dataset 1 contain
approximately 70,000 sequences, whereas dataset 2 contains 4 clusters
with more than 1 million sequences. Dataset 3 (SRR950477) has been
included to benchmark starcode against rainbow in paired-end
clustering mode.
\begin{table}[h]
\centering
\caption{Summary of the biological datasets used for benchmarking.
All the datasets are Illumina reads.}
\begin{tabular}{l|l c r}
    \textit{dataset}   & \textit{read count} &
    \textit{read length} & \textit{type} \\ \hline
    SRR950457 & 6,542,309    & 16     & single     \\
    PRJEB7686 & 127,675,537  & 50     & single     \\
    SRR950477 & 2,460,226    & 50+50  & paired-end \\
\end{tabular}
\label{tab:biobench_properties}
\end{table}

@@

\begin{table}[h]
\centering
\caption{Running time (in seconds) of the software on three biological
datasets. Runs longer than 10 days were interrupted. A dash indicates
that the software cannot be used for this dataset.}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{SRR950457} &
      \textit{PRJEB7686} & \textit{SRR950477} \\ \hline
    starcode   & 5           & 2,898       & 44    \\
    seed       & -           & 60,374      & -     \\
    sidesort   & 4,055       & $>$ 10 days & -     \\
    rainbow    & -           & -           & 306   \\
    cd-hit-est & 170         & 512,591     & -     \\
\end{tabular}
\label{tab:biobench_runtime}
\end{table}

@@

\begin{table}[h]
\centering
\caption{Memory usage (in GB).}
\begin{tabular}{l|c c c}
    \textit{software}   & \textit{SRR950457} &
      \textit{PRJEB7686} & \textit{SRR950477} \\ \hline
    starcode   & 0.65 & 30.9 & 5.2 \\
    seed       & -    & 53.9 & -   \\
    sidesort   & 1.30 & 13.9 & -   \\
    rainbow    & -    & -    & 0.5 \\
    cd-hit-est & 0.80 & 28.5 & -   \\
\end{tabular}
\label{tab:biobench_memory}
\end{table}

@@

\section{Discussion and conclusion}
Starcode is a solid algorithm for sequence clustering based on
all-pairs matching. It achieves high precision, and on experimental
datasets it can be faster than popular heuristics. By design,
starcode is tailored to process high throughput sequencing data
on multi-core platforms with sufficient amount of memory. Due to its
superior precision and faster running time, it fills a gap among
available software, by allowing to take full advantage of middle
to high end hardware.

The speed and precision of starcode also makes it useful for other
clustering tasks, such as identifying enriched motifs in microbial
genomes and in experimental data. Here we have given two examples of such
applications. In the first, we recover a known enriched 12-mer in the
genome of \textit{Neisseria meningitidis}. In the second, we recover
the motif of the human RNA binding protein SRSF1 and notice that it
seems to consist of two halves separated by a linker. This hypothesis
is consistent with the fact that SRSF1 binds RNA through two
consecutive RNA Recognition Motifs (RRM) that are known to bind
3-4 nucleotides in a row\citep{pmid23253355}. The Levenshtein distance,
which incorporates insertions and deletions is more likely to capture
bi-partite binding motifs than position weight matrix representations.
The use of a clustering method to tackle this problem is unusual, but it
illustrates the potential advantages of distance-based approaches.
